---
title: "Applications of CoxianPhaseTypeR"
author: "Dulan Dias et al."
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Applications of CoxianPhaseTypeR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Introduction
This vignette demonstrates how to apply the `CoxianPhaseTypeR` package to both **simulated** and **real-world** survival data, showcasing parameter estimation, model selection, and visualization.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message=FALSE, warning=FALSE)
devtools::install()
library(CoxianPhaseTypeR)
library(ggplot2)
library(survival)
library(dplyr)
library(gridExtra)
```

# Simulated Data Example

## Generate Data
```{r}
set.seed(42)
sim_data <- simulate_coxian(n = 200, lambda = c(2.0, 1.5, 1.2), mu = c(1.0, 0.8, 0.6))
head(sim_data)
```

## Fit Coxian-PH Model
```{r}
fit <- fit_coxian(sim_data, num_phases = 3)
print(fit)
```

## Model Evaluation (AIC, BIC, Log-Likelihood)
```{r}
model_fit <- compute_model_fit(fit, num_params = 6, num_obs = length(sim_data))
print(model_fit)
```

## Determine Optimal Phases
```{r}
best_model <- determine_optimal_phases(sim_data, max_phases = 5)
print(best_model)
```

## Goodness-of-fit statistics
```{r}
compute_goodness_of_fit(fit, sim_data)
```

## Visualization

### Survival Curve: Kaplan-Meier vs Coxian-PH
```{r}
plot_survival_fit(fit, sim_data)
```

### Model Selection Plot: AIC/BIC vs Number of Phases
```{r}
bic_values <- sapply(1:5, function(m) {
  fit_tmp <- fit_coxian(sim_data, num_phases = m)
  compute_model_fit(fit_tmp, num_params = 2 * m, num_obs = length(sim_data))$BIC
})

plot(1:5, bic_values, type = "b", col = "blue", pch = 16, xlab = "Number of Phases",
     ylab = "BIC Value", main = "Model Selection: BIC vs Number of Phases")
```

### Histogram of Simulated Survival Times
```{r}
ggplot(data.frame(time = sim_data), aes(x = time)) +
  geom_histogram(binwidth = 5, fill = "steelblue", color = "black", alpha = 0.7) +
  labs(title = "Histogram of Simulated Survival Times", x = "Time", y = "Frequency") +
  theme_minimal()
```

### State Transition Diagram
```{r}
plot_coxian(fit)
```

# Real-World Data Example (Lung Cancer Dataset)

## Load and Preprocess Data

```{r}
library(survival)
real_data <- lung$time
real_data <- real_data[!is.na(real_data)]

# Summary statistics
summary(real_data)

# Histogram to check distribution
hist(real_data, breaks = 30, main = "Distribution of Lung Cancer Survival Times", xlab = "Survival Time (days)")
```
Since the data contains very large values or outliers, normalize the survival times:
```{r}
real_data <- real_data / max(real_data)
```


## Fit Coxian-PH Model
```{r}
fit_real <- fit_coxian(real_data, num_phases = 3)
print(fit_real)
```

## Model Selection
```{r}
best_model_real <- determine_optimal_phases(real_data, max_phases = 5)
print(best_model_real)
```

## Goodness-of-fit statistics
```{r}
compute_goodness_of_fit(fit_real, real_data)
```

## Visualization

### Kaplan-Meier vs Coxian-PH for Real Data
```{r}
plot_survival_fit(fit_real, real_data)
```

### State Transition Diagram (Real Data)
```{r}
plot_coxian(fit_real)
```

# Summary of Results

## Estimated Parameters
```{r}
df_params <- data.frame(
  Phase = 1:3,
  Lambda = fit$lambda,
  Mu = fit$mu
)
print(df_params)
```

## Model Selection Metrics
```{r}
df_model_fit <- data.frame(
  Model = c("Simulated Data", "Real Data"),
  LogLikelihood = c(model_fit$log_likelihood, best_model_real$best_model$log_likelihood),
  AIC = c(model_fit$AIC, best_model_real$best_model$AIC),
  BIC = c(model_fit$BIC, best_model_real$best_model$BIC)
)
print(df_model_fit)
```

Download Plots
```{r}
ggsave("kaplan_vs_coxian_simulated.png", plot_survival_fit(fit, sim_data))
ggsave("state_transition_simulated.png", plot_coxian(fit))

ggsave("kaplan_vs_coxian_real.png", plot_survival_fit(fit_real, real_data))
ggsave("state_transition_real.png", plot_coxian(fit_real))
```

```{r}
ggsave("log_likelihood_simulated.png", plot_log_likelihood(fit), dpi = 300)
ggsave("log_likelihood_real.png", plot_log_likelihood(fit_real), dpi = 300)
```

